---
title: "Multi-species parameter optimization"
output: rmarkdown::html_vignette
author: cxr team
vignette: >
  %\VignetteIndexEntry{Multi-species parameter optimization}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

**Introduction**

This document explains how to use data obtained from experiments or field observations to predict species fecundity as a function of the intrinsic ability of species to produce offspring ($\lambda$), and interaction coefficients with other species ($\alpha$). We obtain these parameters from our example dataset by fitting increasingly complex population dynamics models of the following form, which is basically a Beverton-Holt model:

$$F = \frac{\lambda}{1+\alpha}$$

Increasingly complex modifications of this model allow us to obtain the interaction matrices for pairs of species, and the effect of covariables (including biotic and abiotic factors) on both $\lambda$ and $\alpha$ coefficients (see vignette `Data and model formats` for details on the five models implemented). The general procedure implemented in this vignette is that the outpupt of a given model is taken as the initial values for the next one (e.g. the `lambda` parameter estimated by model 1 will be the input to model 2, etc.). This procedure is intended to avoid overfitting problems and promote convergence of the more complex models.

All together, the parameters obtained with our model fitting are key for obtaining information of different drivers determining the outcome of species interaction at local scales, which are widely used in "modern" coexistence studies. Most of this vignette presents data preparation from how is obtained in the field to the appropriate format for the analysis. 

**Setup**

We start loading the package and the associated datasets (see vignette `Data and model formats` for details on the data). We will use the following data:

- `competition`: measurement of fruit and seed set of plant individuals in presence of competitors of the same or other species.

- `salinity`: soil salinity for each subplot. This is our abiotic covariate, but any other types of covariates can be included. 

```{r message=F, warning=F}
library(cxr)
data(competition, package = "cxr")
data(salinity, package = "cxr")
```

**Preparing the data**

First we need to format the example dataset for the analysis and add the initial values needed as inputs for model parameterization.

```{r message=F, warning=F}
# spread the data from long to wide format
competition.data <- tidyr::spread(competition,competitor,number,fill = 0)
# how many focal species
focal.sp <- sort(unique(competition.data$focal))
# how many competitor species - in our dataset same set as focal species
competitor.sp <- sort(unique(competition$competitor))
# obtain the competition matrix
comp.matrix <- as.matrix(competition.data[,which(names(competition.data) %in% competitor.sp)]) 

# the rows in the covariates datasets (salinity, in our example) 
# must match the observations of the competition dataset.
# in order to be sure that they match, join them in a single dataset 
# (note that they share a unique identifier, given by plot+subplot+year)
full.data <- dplyr::left_join(competition.data,salinity)

```

Now, we set some general variables regarding the models we want to use, and the parameters we want to estimate. First, the set of models we are using.

```{r message=F, warning=F}
models <- 1:5 
```

Again, fitting modeles sequentially means that the output from model 1 will be taken as initial estimates for model 2, the output from model 2 will be the initial estimates for model 3, and so on. Note, as well, that models 4 and 5 differ from model 3 in that they include the effect of covariates on species seed production in the absence of competition and interactions. Therefore, you should only include these models if you have at least one covariate in your data (see *Lanuza, Bartomeus, Godoy. 2018. Ecology Letters 21:865-874*).

We can estimate the whole parameter set for each model, or fix certain parameters, for example, in case there are independent field estimation of some of them. This information is passed to the function in a list, in which each element is a string vector giving the set of parameters that are to be optimized for each model. In our example, we optimize all parameters (note that model 1 only fits $\lambda$, models 2 and 3 fit $\lambda$ and $\alpha$, whereas models 4 and 5 also include the effect of covariates).

```{r message=F, warning=F}
# which parameters do we optimize for each model?
param.list <- list(m1 = c("lambda"),
                   m2 = c("lambda","alpha"),
                   m3 = c("lambda","alpha"),
                   m4 = c("lambda","alpha","lambda.cov","alpha.cov"),
                   m5 = c("lambda","alpha","lambda.cov","alpha.cov"))
```

In case you had independent estimates of parameters, e.g. $\lambda$, you would remove it from this list, something along these lines:

```{r message=F, warning=F}
#param.list <- list(m2 = c(alpha"),
#                   m3 = c(alpha"),
#                   m4 = c("alpha","lambda.cov","alpha.cov"),
#                   m5 = c("alpha","lambda.cov","alpha.cov")) 
# init.lambda <- your_lambda_estimates
```

Keep the model definitions in a list, for ease:

```{r message=F, warning=F}
fitness.models <- list(BH1 = model_BH1,
                       BH2 = model_BH2,
                       BH3 = model_BH3,
                       BH4 = model_BH4,
                       BH5 = model_BH5)
```

Get back the covariates, after matching with the competition observations. It is important that covariates are passed in a matrix format. If they are named, the output will also include the names.

```{r message=F, warning=F}
# environmental covariates
covariates <- as.matrix(full.data[,"sum_salinity"])
colnames(covariates) <- "sum_salinity"
# if no covariates, comment above and uncomment here
# covariates <- 0
```

Here we specify which optimization methods to use. You can include more than one (but see below). In this example, we use optim_NM because it is the default method in the base `optim` function in R.

```{r message=F, warning=F}
# optimization methods to use
optim.methods <- c("optim_NM" 
                   # "optim_L-BFGS-B"
                  # "nloptr_CRS2_LM"
                   # "nloptr_ISRES"
                   # "nloptr_DIRECT_L_RAND"
                    # "GenSA"
                   # "hydroPSO"
                   # "DEoptimR"
)
```

Since the routine takes the input for a given model from the output of the previous model, we need to specifiy, if we selected more than one optimization method, from which method it will take these initial values. This means that even if we want to try different optimization methods on different models it is wise to run them one by one.

```{r message=F, warning=F}
# from which method are we taking initial estimates for the next model?
init.par.method <- optim.methods[1]
```

We also need to specify the initial values for the first model (model 1, in this example). If you have independent estimates of any parameter, comment out the appropriate lines. The `optim_NM` method used in this example does not account for upper and lower parameter values, but we include them here for completeness.

```{r message=F, warning=F}
# values for initial estimates of parameters. 
# Overwrite if necessary
# init.lambda is calculated after log.fitness for each focal species
lower.lambda <- 1 # minimum number of seeds produced in the absence of competition
upper.lambda <- 1e3
# sigma
lower.sigma <- 1e-5
upper.sigma <- 1
# alpha
init.alpha <- 1e-3
lower.alpha <- 1e-3
upper.alpha <- 1e3
# lambda.cov
init.lambda.cov <- 1 # default to effect of covariates equal to the value of the covariate
lower.lambda.cov <- 1e-10 # allow for very small effects of covariates
upper.lambda.cov <- 1e3
# alpha.cov
init.alpha.cov <- 1
lower.alpha.cov <- 1e-10
upper.alpha.cov <- 1e3
```

Importantly, the package allows computing estimations of standard errors. This error estimation is done via bootstrapping, for two reasons. One is that obtaining estimation of error using Hessian matrices is unstable due to several computational issues, but sencondly and more importantly, error distributions of your variable might not follow a normal distribution, and bootstrapping is a non-parametric method that may recover any error distribution. Please remember that the estimation of errors can be, depending on the model and optimization method, computationally expensive. We recommend 999 runs, yet we provide here a minimal example of it (setting `bootstrap.samples` to only 3 runs).

```{r message=F, warning=F}
generate.errors <- TRUE
bootstrap.samples <- 3
```

And a flag for storing or not the results of the analysis (in this example, actual writing functions are commented out).

```{r message=F, warning=F}
# store results?
write.results <- TRUE
```

**Initialize results structure** 

We will store the results of our analysis in a nested list, following the nested design of the optimization proccess. For each focal species, we calculate the five different models (given in `models`), and for each of these models, we obtain the parameterization from a series of optimization methods (given in `optim.methods`). Thus, we create a list, named `param_estimates`, of the form:

`param_estimates[[species]][[model]][[method]]`

We create this empty placeholder for the outcome with a series of nested loops:

```{r}
param_estimates <- list()
for(i.sp in 1:length(focal.sp)){
  param_estimates[[i.sp]] <- list()
  for(i.model in 1:length(models)){
    param_estimates[[i.sp]][[i.model]] <- list()
    for(i.method in 1:length(optim.methods)){
      param_estimates[[i.sp]][[i.model]][[i.method]] <- list(lambda = NA, 
                                                            lambda.lower.error = NA,
                                                            lambda.upper.error = NA,
                                                            sigma = NA,
                                                            alpha = NA,
                                                            alpha.lower.error = NA,
                                                            alpha.upper.error = NA,
                                                            lambda.cov = NA,
                                                            lambda.cov.lower.error = NA,
                                                            lambda.cov.upper.error = NA,
                                                            alpha.cov = NA,
                                                            alpha.cov.lower.error = NA,
                                                            alpha.cov.upper.error = NA,
                                                            log.likelihood = NA)
    }
    names(param_estimates[[i.sp]][[i.model]]) <- optim.methods
  }
  names(param_estimates[[i.sp]]) <- names(fitness.models)[models]
}
names(param_estimates) <- focal.sp
```

With this setting, we are now ready to optimize the selected parameters based on our dataset. This is done species by species. As the main function, `pm_optim`, accepts the input for a single species, we code a loop for going through all species. Inside this loop, we include the following steps:

1. subset the competition data for the focal species 

2. generate the actual initial values, considering all situations. Remember that, depending on the model, some parameters will have different length. For example, $\alpha$ is a single value in model 2 and a matrix in model 3 onwards. Thus, for species *i*, $\alpha_{ij}$ is a vector of length equal to the number of competitor species.

3. the actual call to `pm_optim`

4. store the results in the nested list we created above

5. update the initial values to be passed on to the next model


```{r message=F, warning=F}
for(i.sp in 1:length(focal.sp)){
  
  # 1) subset and prepare the data
  
  focal.sp.data <- subset(competition.data, focal == focal.sp[i.sp])
  # current focal species
  focal <- unique(focal.sp.data$focal)

  # we may have observations in which the focal individual produced no seeds. 
  # The optimization methods work with the logarithm of seed number,
  # so add a small offset to prevent numerical errors.
  focal.sp.data$seed[focal.sp.data$seed == 0] <- 0.001
  # our fitness metric is simply number of produced seeds
  fitness <- focal.sp.data$seed 
  log.fitness <- log(fitness)
  # competition matrix: number of competitors
  focal.comp.matrix <- comp.matrix[which(competition.data$focal == focal.sp[i.sp]),]
  # number of competitors
  num.competitors <- dim(focal.comp.matrix)[2]
  # number of covariates
  num.covariates <- ifelse(is.null(ncol(covariates)),0,ncol(covariates))
  # covariates for the focal species
  if(num.covariates > 0){
    focal.covariates <- covariates[which(competition.data$focal == focal.sp[i.sp]),,drop = FALSE]
  }else{
    focal.covariates <- 0
  }
  
  # 2) generate initial values for the different parameters
  # or gather them from data if they are not to be optimized
  
  # lambda
  if("lambda" %in% unlist(param.list)){
    current.init.lambda <- mean(log.fitness)
  }else{
    current.init.lambda <- init.lambda[i.sp]
  }
  # sigma
  current.init.sigma <- sd(log.fitness)
  if(current.init.sigma > upper.sigma){
    current.init.sigma <- upper.sigma
  }
  # alpha
  if("alpha" %in% unlist(param.list)){
    if(models[1]<=2){
      alpha.length <- 1
    }else{
      alpha.length <- num.competitors
    }
    if(length(init.alpha) != alpha.length){
     current.init.alpha <- rep(init.alpha[1],alpha.length) 
    }else{
      current.init.alpha <- init.alpha
    }
  }else{
    current.init.alpha <- init.alpha[i.sp,]
  }
  # lambda.cov
  if("lambda.cov" %in% unlist(param.list)){
    if(length(init.lambda.cov) != num.covariates){
      current.init.lambda.cov <- rep(init.lambda.cov[1],num.covariates)
    }else{
      current.init.lambda.cov <- init.lambda.cov  
    }
  }else{
    current.init.lambda.cov <- init.lambda.cov[i.sp]  
  }
  # alpha.cov
  if("alpha.cov" %in% unlist(param.list)){
    if(models[1]<=4){
      length.alpha.cov <- num.covariates
    }else if(models[1]>4){
      length.alpha.cov <- num.covariates*num.competitors
    }
    if(length(init.alpha.cov) != length.alpha.cov){
      current.init.alpha.cov <- rep(init.alpha.cov[1],length.alpha.cov)
    }else{
      current.init.alpha.cov <- init.alpha.cov  
    }  
  }else{
    current.init.alpha.cov <- init.alpha.cov[i.sp]  
  }
  
  # model to optimize  
  for(i.model in 1:length(models)){
    
  # 3) compute each method
    
    for(i.method in 1:length(optim.methods)){
      
      temp.results <- pm_optim(fitness.model = fitness.models[[models[i.model]]],
                                   optim.method = optim.methods[i.method],
                                   param.list = param.list[[i.model]],
                                   log.fitness = log.fitness,
                                   init.lambda = current.init.lambda,
                                   lower.lambda = lower.lambda,
                                   upper.lambda = upper.lambda,
                                   init.sigma = current.init.sigma,
                                   lower.sigma = lower.sigma,
                                   upper.sigma = upper.sigma,
                                   init.alpha = current.init.alpha,
                                   lower.alpha = lower.alpha,
                                   upper.alpha = upper.alpha,
                                   init.lambda.cov = current.init.lambda.cov,
                                   lower.lambda.cov = lower.lambda.cov,
                                   upper.lambda.cov = upper.lambda.cov,
                                   init.alpha.cov = current.init.alpha.cov,
                                   lower.alpha.cov = lower.alpha.cov,
                                   upper.alpha.cov = upper.alpha.cov,
                                   focal.comp.matrix = focal.comp.matrix,
                                   focal.covariates = focal.covariates,
                                   generate.errors = generate.errors,
                                   bootstrap.samples = bootstrap.samples,verbose = TRUE)
      ###############
      # 4) store results
      
      param_estimates[[i.sp]][[i.model]][[i.method]]$lambda <- 
        temp.results$lambda
      
      param_estimates[[i.sp]][[i.model]][[i.method]]$lambda.lower.error <- 
        temp.results$lambda.lower.error
      
      param_estimates[[i.sp]][[i.model]][[i.method]]$lambda.upper.error <- 
        temp.results$lambda.upper.error
      
      param_estimates[[i.sp]][[i.model]][[i.method]]$sigma <- 
        temp.results$sigma
      
      param_estimates[[i.sp]][[i.model]][[i.method]]$alpha <- 
        temp.results$alpha
      
      param_estimates[[i.sp]][[i.model]][[i.method]]$alpha.upper.error <- 
        temp.results$alpha.upper.error
      
      param_estimates[[i.sp]][[i.model]][[i.method]]$alpha.lower.error <- 
        temp.results$alpha.lower.error
      
      param_estimates[[i.sp]][[i.model]][[i.method]]$lambda.cov <- 
        temp.results$lambda.cov
      
      param_estimates[[i.sp]][[i.model]][[i.method]]$lambda.cov.upper.error <- 
        temp.results$lambda.cov.upper.error
      
      param_estimates[[i.sp]][[i.model]][[i.method]]$lambda.cov.lower.error <- 
        temp.results$lambda.cov.lower.error
      
      param_estimates[[i.sp]][[i.model]][[i.method]]$alpha.cov <- 
        temp.results$alpha.cov
      
      param_estimates[[i.sp]][[i.model]][[i.method]]$alpha.cov.upper.error <- 
        temp.results$alpha.cov.upper.error
      
      param_estimates[[i.sp]][[i.model]][[i.method]]$alpha.cov.lower.error <- 
        temp.results$alpha.cov.lower.error
      
      param_estimates[[i.sp]][[i.model]][[i.method]]$log.likelihood <- 
        temp.results$log.likelihood
      
    }# for i.method
    
    #######################
    # 5) update initial values for the next model, if there is one
    
    # lambda
    if(i.model < length(param.list)){
      if("lambda" %in% param.list[[i.model+1]]){
        if(!is.na(param_estimates[[i.sp]][[i.model]][[init.par.method]]$lambda)){
          current.init.lambda <- param_estimates[[i.sp]][[i.model]][[init.par.method]]$lambda
        }
      }
      # sigma
      if(!is.na(param_estimates[[i.sp]][[i.model]][[init.par.method]]$sigma)){
        current.init.sigma <- param_estimates[[i.sp]][[i.model]][[init.par.method]]$sigma
        if(current.init.sigma > upper.sigma){
          current.init.sigma <- upper.sigma
        }
      }
      # alpha
      if("alpha" %in% param.list[[i.model+1]]){
        if(sum(is.na(param_estimates[[i.sp]][[i.model]][[init.par.method]]$alpha)) == 0){
          current.init.alpha <- param_estimates[[i.sp]][[i.model]][[init.par.method]]$alpha
          # is the current estimate of the appropriate length?
          if(models[i.model+1] > 2){
            if(length(current.init.alpha) == 1){
              current.init.alpha <- rep(current.init.alpha,num.competitors)
            }
          }# if model > 2
        }
      }
      # lambda.cov
      if("lambda.cov" %in% param.list[[i.model+1]]){
        if(sum(is.na(param_estimates[[i.sp]][[i.model]][[init.par.method]]$lambda.cov)) == 0){
          current.init.lambda.cov <- 
            param_estimates[[i.sp]][[i.model]][[init.par.method]]$lambda.cov
        }
      }
      # alpha.cov
      if("alpha.cov" %in% param.list[[i.model+1]]){
        if(sum(is.na(param_estimates[[i.sp]][[i.model]][[init.par.method]]$alpha.cov)) == 0){
          current.init.alpha.cov <- 
            param_estimates[[i.sp]][[i.model]][[init.par.method]]$alpha.cov
          # is the current estimate of the appropriate length?
          if(models[i.model+1] > 4){
            if(length(current.init.alpha.cov) == num.covariates){
              current.init.alpha.cov <- rep(current.init.alpha.cov,num.competitors)
            }
          }# if model > 4
        }
      }# alpha.cov
    }# if still models to run
  }# for i.model
}# for i.sp

```

If we set the `write.results` flag, the nested list will be written to disk as a .RData file. The list structure is very convenient for handling these results in R, as it maintains the nested design of the parameterization (per species, model, and optimization method). However, it may also be necessary to handle the results as dataframes and csv files, e.g. for plotting. Here we build four dataframes, in long format, for the different sets of parameters: lambda, alpha, lambda.cov, and alpha.cov. This code assumes that covariates are named, so that vectors returned from the list are also named.

```{r message=F, warning=F}
if(write.results){
  # save(param_estimates,file = "./results/param_estimates.RData")
  
  # just in case it is a factor
  focal.sp <- sort(as.character(focal.sp))
  competitors <- 
    sort(names(
      param_estimates[[1]][[names(fitness.models)[models[models == max(models)]]]][[1]]$alpha))
  my.models <- names(fitness.models)[models]
  my.covariates <- c("sum_salinity")
  
  # lambda
  lambda.values <- expand.grid(focal.sp,my.models,optim.methods,
                               stringsAsFactors = F)
  names(lambda.values) <- c("species","model","method")
  lambda.values$lambda <- NA_real_
  lambda.values$lambda.lower <- NA_real_
  lambda.values$lambda.upper <- NA_real_
  
  # alpha
  alpha.values <- expand.grid(focal.sp,competitors,my.models,optim.methods,
                              stringsAsFactors = F)
  names(alpha.values) <- c("focal","competitor","model","method")
  alpha.values$alpha <- NA_real_
  alpha.values$alpha.lower <- NA_real_
  alpha.values$alpha.upper <- NA_real_
  
  # lambda.cov
  lambda.cov.values <- expand.grid(focal.sp,my.models,optim.methods,my.covariates,
                                   stringsAsFactors = F)
  names(lambda.cov.values) <- c("species","model","method","covariate")
  lambda.cov.values$lambda.cov <- NA_real_
  lambda.cov.values$lambda.cov.lower <- NA_real_
  lambda.cov.values$lambda.cov.upper <- NA_real_
  
  # alpha.cov
  alpha.cov.values <- expand.grid(focal.sp,competitors,my.models,optim.methods,my.covariates,
                                  stringsAsFactors = F)
  names(alpha.cov.values) <- c("focal","competitor","model","method","covariate")
  alpha.cov.values$alpha.cov <- NA_real_
  alpha.cov.values$alpha.cov.lower <- NA_real_
  alpha.cov.values$alpha.cov.upper <- NA_real_
  
  # fill up the dataframes
  for(i.sp in 1:length(focal.sp)){
    for(i.model in 1:length(my.models)){
      for(i.method in 1:length(optim.methods)){
        
        # lambda
        lambda.pos <- which(lambda.values$species == focal.sp[i.sp] &
                              lambda.values$model == my.models[i.model] &
                              lambda.values$method == optim.methods[i.method])
        lambda.values$lambda[lambda.pos] <- 
          param_estimates[[focal.sp[i.sp]]][[my.models[i.model]]][[optim.methods[i.method]]]$lambda
        
        lambda.values$lambda.lower[lambda.pos] <- 
          param_estimates[[focal.sp[i.sp]]][[my.models[i.model]]][[optim.methods[i.method]]]$lambda.lower.error
        
        lambda.values$lambda.upper[lambda.pos] <- 
          param_estimates[[focal.sp[i.sp]]][[my.models[i.model]]][[optim.methods[i.method]]]$lambda.upper.error
        
        # alpha
        my.alpha.vector <- 
          param_estimates[[focal.sp[i.sp]]][[my.models[i.model]]][[optim.methods[i.method]]]$alpha
        
        my.alpha.lower.vector <- 
          param_estimates[[focal.sp[i.sp]]][[my.models[i.model]]][[optim.methods[i.method]]]$alpha.lower.error
        
        my.alpha.upper.vector <- 
          param_estimates[[focal.sp[i.sp]]][[my.models[i.model]]][[optim.methods[i.method]]]$alpha.upper.error
        
        if(length(my.alpha.vector)>1){ # model 3 onwards
          alpha.pos <- which(alpha.values$focal == focal.sp[i.sp] &
                               alpha.values$model == my.models[i.model] &
                               alpha.values$method == optim.methods[i.method] &
                               alpha.values$competitor %in% names(my.alpha.vector))
        }else{ # model 2, i.e. a single alpha for all competitors
          alpha.pos <- which(alpha.values$focal == focal.sp[i.sp] &
                               alpha.values$model == my.models[i.model] &
                               alpha.values$method == optim.methods[i.method])
        }
        
        alpha.values$alpha[alpha.pos] <- my.alpha.vector
        alpha.values$alpha.lower[alpha.pos] <- my.alpha.lower.vector
        alpha.values$alpha.upper[alpha.pos] <- my.alpha.upper.vector
        
        # lambda.cov
        my.lambda.cov.vector <- 
          param_estimates[[focal.sp[i.sp]]][[my.models[i.model]]][[optim.methods[i.method]]]$lambda.cov
        
        my.lambda.cov.lower.vector <- 
          param_estimates[[focal.sp[i.sp]]][[my.models[i.model]]][[optim.methods[i.method]]]$lambda.cov.lower.error
        
        my.lambda.cov.upper.vector <- 
          param_estimates[[focal.sp[i.sp]]][[my.models[i.model]]][[optim.methods[i.method]]]$lambda.cov.upper.error
        
        # all covariates at once, should be ok
        lambda.cov.pos <- which(lambda.cov.values$species == focal.sp[i.sp] &
                                  lambda.cov.values$model == my.models[i.model] &
                                  lambda.cov.values$method == optim.methods[i.method])
        
        lambda.cov.values$lambda.cov[lambda.cov.pos] <- my.lambda.cov.vector
        lambda.cov.values$lambda.cov.lower[lambda.cov.pos] <- my.lambda.cov.lower.vector
        lambda.cov.values$lambda.cov.upper[lambda.cov.pos] <- my.lambda.cov.upper.vector
        
        # alpha.cov
        my.alpha.cov.vector <- 
          param_estimates[[focal.sp[i.sp]]][[my.models[i.model]]][[optim.methods[i.method]]]$alpha.cov
        
        my.alpha.cov.lower.vector <- 
          param_estimates[[focal.sp[i.sp]]][[my.models[i.model]]][[optim.methods[i.method]]]$alpha.cov.lower.error
        
        my.alpha.cov.upper.vector <- 
          param_estimates[[focal.sp[i.sp]]][[my.models[i.model]]][[optim.methods[i.method]]]$alpha.cov.upper.error
        
        # differentiate models 4 and 5
        if(!is.null(my.covariates) & !is.na(my.alpha.cov.vector[1])){
          if(length(my.alpha.cov.vector) == length(my.covariates)){
            my.alpha.cov.comp <- competitors
          }else{
            my.alpha.cov.comp <- substr(names(my.alpha.cov.vector),
                                        stringr::str_length(names(my.alpha.cov.vector))-3,
                                        stringr::str_length(names(my.alpha.cov.vector)))
          }# if-else single alpha.cov
          
          for(i.covariate in 1:length(my.covariates)){
            my.cov <- 
              my.alpha.cov.vector[which(grepl(my.covariates[i.covariate],names(my.alpha.cov.vector)))]
            
            my.lower.cov <- 
              my.alpha.cov.lower.vector[which(grepl(my.covariates[i.covariate],names(my.alpha.cov.lower.vector)))]
            
            my.upper.cov <- 
              my.alpha.cov.upper.vector[which(grepl(my.covariates[i.covariate],names(my.alpha.cov.upper.vector)))]
            
            alpha.cov.pos <- which(alpha.cov.values$focal == focal.sp[i.sp] &
                                     alpha.cov.values$model == my.models[i.model] &
                                     alpha.cov.values$method == optim.methods[i.method] &
                                     alpha.cov.values$competitor %in% my.alpha.cov.comp & 
                                     alpha.cov.values$covariate == my.covariates[i.covariate])
            
            alpha.cov.values$alpha.cov[alpha.cov.pos] <- my.cov
            alpha.cov.values$alpha.cov.lower[alpha.cov.pos] <- my.lower.cov
            alpha.cov.values$alpha.cov.upper[alpha.cov.pos] <- my.upper.cov
          }
        }# if covariates and alpha.cov.values
        
      }# for each method
    }# for each model
  }# for each sp
  
  # write.csv2(lambda.values,file = "your/path/lambda_values.csv",row.names = F,append = F)
  # write.csv2(alpha.values,file = "your/path/alpha_values.csv",row.names = F,append = F)
  # write.csv2(lambda.cov.values,file = "your/path/lambda_cov_values.csv",row.names = F,append = F)
  # write.csv2(alpha.cov.values,file = "your/path/alpha_cov_values.csv",row.names = F,append = F)

}

```

Each combination of focal species, model, and optimization method has an estimated $\lambda$

```{r}
  head(lambda.values)
```

Likewise for interaction coefficients, $\alpha$

```{r}
  head(alpha.values[alpha.values$model == "BH2",]) # first model to fit alpha
```

And for $\Theta$, or `lambda.cov` in the models, representing the effect of covariates over $\lambda$

```{r}
  head(lambda.cov.values[lambda.cov.values$model == "BH4",]) # first model to fit lambda.cov
```

Lastly, $\phi$ or `alpha.cov` represents the effect of covariates over $\alpha$

```{r}
  head(alpha.cov.values[alpha.cov.values$model == "BH4",]) # first model to fit alpha.cov
```


