---
title: "Getting started with cxr"
output: rmarkdown::html_vignette
author: cxr team
vignette: >
  %\VignetteIndexEntry{Getting started}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

**Introduction**

The `cxr` package provides a general interface to parameterize population dynamics models from empirical data, and obtain from the estimations of these parameters (species vital rates and interaction coefficients) a series of metrics associated with modern coexistence theory. These metrics measure species competitive ability, their responses to competition as well as niche and average fitness differences (both demographic and competitive response differences).

Here we demonstrate the basic usage of the package, using the data included in it (see *Lanuza, Bartomeus, Godoy. 2018. Ecology Letters 21:865-874* for a description of the dataset). In particular, with this dataset, we will obtain the values of seed production in the absence of neighbours and the strength of competitive interactions between pairs of species. These values are the basis for estimating the degree of niche overlap and average fitness differences between any pair of species. 

**Set up**

First, we load the package and the associated data

```{r}
library(cxr)
data(competition, package = "cxr")
```

The `competition` dataset includes observation from *20* different plant species (see vignette `Data and model formats` for details on the included datasets). We will subset the information to consider only two of them, *Chamaemelum fuscatum* and *Leontodon maroccanus*, which are coded as *CHFU* and *LEMA* respectively.

```{r message=F, warning=F}
my.sp <- c("CHFU", "LEMA")
two.sp.data <- subset(competition, focal %in% my.sp & competitor %in% my.sp)
```

In order to estimate the different parameters for both species, we need to store separately the information on their observations as focal species. For this, we transform the data from long to wide format, and obtain the competition matrices for each focal species (where each row is an observation of a focal individual, and each column one species).

```{r message=F, warning=F}
# transform from long to wide format
two.sp.wide <- tidyr::spread(data = two.sp.data,key = competitor, value = number)
# get competition matrices. First, the complete one
comp.matrix <- as.matrix(two.sp.wide[,which(names(two.sp.wide) %in% my.sp)])

# then subset for each focal species
sp1.comp.matrix <- comp.matrix[which(two.sp.wide$focal == my.sp[1]),]
sp2.comp.matrix <- comp.matrix[which(two.sp.wide$focal == my.sp[2]),]

```

We need to retrieve, as well, the estimation of fitness associated with each observation. In our dataset this is the seed production of focal individuals, for which the number of neighbours (`comp.matrix`) were also taken from field observations. Note that we pass the log of the seed production (fitness metric) for facilitating the optimization procedure.

```{r message=F, warning=F}
sp1.fitness <- two.sp.wide$seed[which(two.sp.wide$focal == my.sp[1])]
sp1.log.fitness <- log(sp1.fitness)

sp2.fitness <- two.sp.wide$seed[which(two.sp.wide$focal == my.sp[2])]
sp2.log.fitness <- log(sp2.fitness)
```

Finally, as with any optimization procedure, we need to provide initial estimates for the parameters we want to fit numerically. In this case, the initial estimate is the mean of the log number of seeds produced.

```{r message=F, warning=F}
init.lambda <- c(mean(sp1.log.fitness),mean(sp2.log.fitness))
```

In `cxr` we have included five different population dynamic models for estimating coexistence parameters. These are increasingly complex variations of a general Beverton-Holt model, and are included in the package as functions `BH_1` to `BH_5` (see vignette `Data and model formats` for further information on these models). We will fit the parameters of model 3. This model differentiates between intra- and inter-specific competition but does not include yet the effects of environmental covariates or other type of trophic interactions besides competition (see *Lanuza et al. 2018*, *Godoy and Levine. 2014. Ecology 95:726-736* for further details).

$$F_i = \frac{\lambda_i}{1+\sum \alpha_{ij}N_j}$$

Specifying an optimization method, we can now call the main function with sensible values of upper and lower parameter bounds. We have implemented optimization with eight different optimization methods, that cover a range of methodologies (see e.g. the help of `cxr_optimize` for the available methods, and *Mullen, K.M. 2014. Journal of Statistical Software 60:1–45* for an overview of optimization methods). Note that not all methods accept lower and upper boundaries, e.g. the standard method `optim_NM` does not, whereas `optim_L-BFGS-B` does accept boundaries. However, `optim_L-BFGS-B` does not allow parameter values for which the optimization function is undefined, and this will return an error. In this simple example, we use this latter method with allowed lower and upper boundaries.  

```{r message=F, warning=F}
sp1.par <- cxr_optimize(fitness.model = BH_3,
                        optim.method = "optim_L-BFGS-B",
                        param.list = c("lambda","alpha"),
                        log.fitness = sp1.log.fitness,
                        init.lambda = init.lambda[1],
                        # negative numbers of seed production are not 
                        # biologically meaningful, 
                        # hence we fix lambda to a minimum of 0.
                        lower.lambda = 0, 
                        upper.lambda = 1e3,
                        init.sigma = sd(sp1.log.fitness),
                        lower.sigma = 1e-5,
                        upper.sigma = 1,
                        # Initial alpha coefficients set to almost no competition. 
                        # In this example, assume alpha > 0 
                        # (alpha <= 0 will throw an error when used 
                        # in combination with optim_L-BFGS-B)
                        init.alpha = c(1e-3,1e-3),
                        lower.alpha = c(1e-3,1e-3),
                        upper.alpha = c(1e3,1e3),
                        focal.comp.matrix = sp1.comp.matrix)

sp2.par <- cxr_optimize(fitness.model = BH_3,
                        optim.method = "optim_L-BFGS-B",
                        param.list = c("lambda","alpha"),
                        log.fitness = sp2.log.fitness,
                        init.lambda = init.lambda[2],
                        lower.lambda = 0,
                        upper.lambda = 1e3,
                        init.sigma = sd(sp2.log.fitness),
                        lower.sigma = 1e-3,
                        upper.sigma = 1,
                        init.alpha = c(1e-3,1e-3),
                        lower.alpha = c(1e-3,1e-3),
                        upper.alpha = c(1e3,1e3),
                        focal.comp.matrix = sp2.comp.matrix)

```

The `cxr_optimize` function returns a list with several components 

```{r message=F, warning=F}
sp1.par
```

In this basic example, we are only interested in obtaining from the optimization process the seed production in the absence of competition (lambda), and the intra- and interspecific interactions, i.e. the row of the interaction matrix that corresponds to our focal species. Other components are NA either because the third model does not account for them (e.g. `lambda.cov` or `alpha.cov`), or because we are not estimating them (e.g. standard errors). In any case, we can retrieve the estimated 2x2 interaction matrix from the results.

```{r message=F, warning=F}
interaction.matrix <- matrix(c(sp1.par$alpha, sp2.par$alpha), 
                             nrow = length(my.sp), 
                             byrow = TRUE)
colnames(interaction.matrix) <- c("CHFU", "LEMA")
rownames(interaction.matrix) <- c("CHFU", "LEMA")
interaction.matrix
```

and obtain the niche overlap and average fitness ratio between the two species. These metrics allow us to understand the expected coexistence between both species given the data (see the help page of the functions for further information and references).

```{r message=F, warning=F}

NicheOverlap(pair.matrix = interaction.matrix)

AvgFitnessRatio(lambda = c(sp1.par$lambda,sp2.par$lambda),pair.matrix = interaction.matrix)

```








