---
title: "Obtain competitive responses and effects"
output: rmarkdown::html_vignette
author: cxr team
vignette: >
  %\VignetteIndexEntry{Competitive responses/effects}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

**Introduction**

In other vignettes we have estimated a series of interactions coefficients between pairs of species. These estimations allow species to be differentiated in their niche and their average fitness (via competitive response differences). However, it is also possible to obtain a metric per species of the average interaction effect to all other species, or their average response when interacting with all other species. This is done by estimating all these coefficients for all species at the same time, and not each species as a time as previous modelling procedures. With this procedure, we do not allow species to be differentiated in their niche (see *Godoy, Kraft, and Levine, 2014, Ecology Letters 17:836-844* for further details). Again, here we consider the plant species from the associated data. Please note that here per-germinant fecundity estimates ($\lambda$) generated independently are used as fixed parameters.

**Set up**

```{r}
library(cxr)
# for usign the pipe operator, %>%
library(magrittr)
data(competition)
data(param_estimates)
```

As in other vignettes, we need to specify which optimization method are we using, and whether we want to generate standard errors from bootstrapping.

```{r}
optim.methods <- c(#"optim_NM"
                   "optim_L-BFGS-B"
                   # "nloptr_CRS2_LM", 
                   # "nloptr_ISRES", 
                   # "nloptr_DIRECT_L_RAND", 
                   # "GenSA"
                   # "hydroPSO", 
                   # "DEoptimR"
)

# if we want quick calculations, we can disable 
# the bootstrapping for the standard errors
generate.errors <- FALSE
bootstrap.samples <- 99 # discarded if false
```

Since we have independent estimates for $\lambda$, we are taking it as a fixed parameter that we do not want to fit. This can be passed on to the \code{ER_optimize} function by choosing the appropriate effect/response model. Note that, if we are not interested in fitting $\lambda$, we can dispense with specifying lower/upper bounds for it.

```{r}
optimize.lambda <- FALSE
# model is different...
if(optimize.lambda){
  effect.response.model <- EffectResponse_lambda
  lower.lambda <- 0
  upper.lambda <- 1e3
}else{
  effect.response.model <- EffectResponse
  lower.lambda <- NA
  upper.lambda <- NA
}
```

Finally, a flag for writing results or not

```{r}
write.results <- FALSE
```

It is possible to estimate, alongside competitive effects and responses, the effect of covariates over these terms. For simplicity, we will assume no effect of covariates in this example, but their addition is similar to the workflow exemplified in the vignette \code{Multi-species parameter optimization}. The dataset included with the package includes observations from 2016, but here we provide general code for any number of years (or sampling campaigns). We group observations in year x site records, where a site is a subplot within a specific plot (e.g. subplot A1 of plot 1).

```{r}
sp.data <- competition %>% dplyr::group_by(year,plot,subplot,focal,competitor) %>% dplyr::summarise(seed = sum(seed),number = sum(number))
names(sp.data)[which(names(sp.data) == "seed")] <- "fitness"
sp.data$site <- paste(sp.data$year,sp.data$plot,sp.data$subplot,sep="_")
sp.data <- sp.data[,c("site","focal","fitness","competitor","number")]

# this is the set of species we are fitting
# only those with proper lambda estimates
focal.sp <- levels(sp.data$focal)
sp.with.lambda <- names(param_estimates)
sp.names <- sort(focal.sp[which(focal.sp %in% sp.with.lambda)])
```

For the \code{ER_optimize} function to work properly, the dataset must be complete, with explicit zeros if a combination focal.sp-competitor was not observed in a given record. In case the sp.data dataframe does not include explicit missing competitors, here is a somewhat convoluted way for setting zeros to it so that for each focal sp, all competitor sp are included. 

```{r}
sites <- unique(sp.data$site)

missing.data <- sp.data
missing.data$site <- "0"
missing.data$focal <- "0"
missing.data$fitness <- 0
missing.data$competitor <- "0"
missing.data$number <- 0
# just in case it overflows
missing.data <- rbind(missing.data,missing.data,missing.data)
count <- 1

for(i.site in 1:length(sites)){
  for(i.sp in 1:length(sp.names)){
    my.competitors <- unique(sp.data$competitor[sp.data$site == sites[i.site] & sp.data$focal == sp.names[i.sp]])
    if(length(my.competitors) > 0 & length(my.competitors) < length(sp.names)){
      my.fitness <- sp.data$fitness[sp.data$site == sites[i.site] & sp.data$focal == sp.names[i.sp]][1]
      
      missing.competitors <- sp.names[which(!sp.names %in% my.competitors)]
      for(i.com in 1:length(missing.competitors)){
        
        missing.data$site[count] <- sites[i.site]
        missing.data$focal[count] <- sp.names[i.sp]
        missing.data$fitness[count] <- my.fitness
        missing.data$competitor[count] <- missing.competitors[i.com]
        missing.data$number[count] <- 0
        
        count <- count + 1
      }# for each missing
    }# if any missing
  }# for i.sp
}# for i.site

missing.data <- droplevels(subset(missing.data,competitor != "0"))

sp.data <- rbind(sp.data,missing.data)
sp.data <- dplyr::arrange(sp.data, focal, site, competitor)
```

Discard focal species with no seeds

```{r}
sp.data <- droplevels(subset(sp.data, fitness > 0))
```

Get independent $\lambda$ values and append them to a dataframe

```{r}
lambda.values <- data.frame(sp = sp.names,lambda = 0, sigma = 0)

# from which model and optimization method are we taking estimates?
estimates.model <- "BH_5" #Beverton-holt model number 5
estimates.method <- "optim_L-BFGS-B"

# gather lambda from fitted data
for(i.sp in 1:length(sp.names)){
  if(!is.null(param_estimates[[sp.names[i.sp]]])){
    lambda.values$lambda[lambda.values$sp == sp.names[i.sp]] <- param_estimates[[sp.names[i.sp]]][[estimates.model]][[estimates.method]]$lambda
    lambda.values$sigma[lambda.values$sp == sp.names[i.sp]] <- param_estimates[[sp.names[i.sp]]][[estimates.model]][[estimates.method]]$sigma
  }
}

# sanity check
lambda.values <- dplyr::arrange(subset(lambda.values, sp %in% sp.data$focal),sp.names)
# sigma is also a parameter
sigma <- mean(lambda.values$sigma)
```

Double-check that we only include species that have proper estimates

```{r}
sp.data <- subset(sp.data, focal %in% lambda.values$sp & competitor %in% lambda.values$sp)
```

Finally, set initial values for e (i.e. competitive effects) and r (i.e competitive responses), as well as their lower and upper bounds. Lower and upper bounds are most likely to be the same for all species, and in that case it is convenient to provide a single value. If we want to specify different values for every species, we can pass vectors of the same length as the lambda, e, and r parameters.

```{r}
r.values <- rep(1,nrow(lambda.values))
e.values <- rep(1,nrow(lambda.values))

lower.r <- 0
upper.r <- 1e2
lower.e <- lower.r
upper.e <- upper.r
lower.sigma <- 0.0000000001
upper.sigma <- 1
```

We can now initialize the results data structure, a list in which each component will consist of the estimates from an optimization method.

```{r}
ER.list <- list()
for(i.method in 1:length(optim.methods)){
  ER.list[[i.method]] <- list(lambda = NA,
                                 lambda.lower.error = NA,
                                 lambda.upper.error = NA,
                                 response = NA,
                                 response.lower.error = NA,
                                 response.upper.error = NA,
                                 effect = NA,
                                 effect.lower.error = NA,
                                 effect.upper.error = NA,
                                 sigma = NA,
                                 lambda.cov = NA,
                                 lambda.cov.lower.error = NA,
                                 lambda.cov.upper.error = NA,
                                 response.cov = NA,
                                 response.cov.lower.error = NA,
                                 response.cov.upper.error = NA,
                                 effect.cov = NA,
                                 effect.cov.lower.error = NA,
                                 effect.cov.upper.error = NA,
                                 log.likelihood = NA)

}
names(ER.list) <- optim.methods
```

In the main loop, we 1) call the \code{ER_optimize} function with each of the selected optimization methods, 2) store the resulting values in the list we just created. Note that, in the function call, we omit the arguments related to the inclusion of covariates, because these have a default \code{NULL} value. These arguments (e.g. \code{lambda.cov}, \code{r.cov}, \code{e.cov}, and their associated errors) are nevertheless returned as NA from the function, and for completeness we also retrieve them.

```{r}
for(i.method in 1:length(optim.methods)){
  
  # 1) call optimization function
  param.results <- ER_optimize(lambda.vector = lambda.values$lambda,
                               e.vector = e.values,
                               r.vector = r.values,
                               sigma = sigma,
                               lower.lambda = lower.lambda,
                               upper.lambda = upper.lambda,
                               lower.e = lower.e,
                               upper.e = upper.e,
                               lower.r = lower.r,
                               upper.r = upper.r,
                               lower.sigma = lower.sigma,
                               upper.sigma = upper.sigma,
                               effect.response.model = effect.response.model,
                               optim.method = optim.methods[i.method],
                               sp.data = sp.data,
                               optimize.lambda = optimize.lambda,
                               generate.errors = generate.errors,
                               bootstrap.samples = bootstrap.samples)
  
  # 2) store results
  ER.list[[optim.methods[i.method]]]$lambda <- param.results$lambda
  ER.list[[optim.methods[i.method]]]$lambda.lower.error <- param.results$lambda.lower.error
  ER.list[[optim.methods[i.method]]]$lambda.upper.error <- param.results$lambda.upper.error
  ER.list[[optim.methods[i.method]]]$response <- param.results$response
  ER.list[[optim.methods[i.method]]]$response.lower.error <- param.results$response.lower.error
  ER.list[[optim.methods[i.method]]]$response.upper.error <- param.results$response.upper.error
  ER.list[[optim.methods[i.method]]]$effect <- param.results$effect
  ER.list[[optim.methods[i.method]]]$effect.lower.error <- param.results$effect.lower.error
  ER.list[[optim.methods[i.method]]]$effect.upper.error <- param.results$effect.upper.error
  ER.list[[optim.methods[i.method]]]$sigma <- param.results$sigma
  ER.list[[optim.methods[i.method]]]$lambda.cov <- param.results$lambda.cov
  ER.list[[optim.methods[i.method]]]$lambda.cov.lower.error <- param.results$lambda.cov.lower.error
  ER.list[[optim.methods[i.method]]]$lambda.cov.upper.error <- param.results$lambda.cov.upper.error
  ER.list[[optim.methods[i.method]]]$response.cov <- param.results$response.cov
  ER.list[[optim.methods[i.method]]]$response.cov.lower.error <- param.results$response.cov.lower.error
  ER.list[[optim.methods[i.method]]]$response.cov.upper.error <- param.results$response.cov.upper.error
  ER.list[[optim.methods[i.method]]]$effect.cov <- param.results$effect.cov
  ER.list[[optim.methods[i.method]]]$effect.cov.lower.error <- param.results$effect.cov.lower.error
  ER.list[[optim.methods[i.method]]]$effect.cov.upper.error <- param.results$effect.cov.upper.error
  ER.list[[optim.methods[i.method]]]$log.likelihood <- param.results$log.likelihood
}

if(write.results){
  save(ER.list,file = "./results/effect_response_estimates.Rdata")
}

```

Next, a simple plot showing the variations in competitive effects and responses among species. For that, we have to format the results into a dataframe

```{r}
er.data <- data.frame(species = sp.names, competitive.effect = ER.list[[1]]$effect, competitive.response = ER.list[[1]]$response)

er.data.long <- tidyr::gather(er.data,key = "var",value = "value",-species)

er.plot <- ggplot2::ggplot(er.data.long,ggplot2::aes(y = species, x = value, group = var))+
  ggplot2::geom_point()+
  ggplot2::facet_grid(.~var, scales = "free_x")+
  NULL
er.plot
```

