---
title: "Getting started with cxr"
output: rmarkdown::html_vignette
author: David García-Callejas and cxr team
vignette: >
  %\VignetteIndexEntry{Getting started}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup,echo=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

**Introduction**

The `cxr` package provides a general interface to obtain estimates of species vital rates and interaction coefficients between species pairs from empirical data. These estimations are critical to parameterize population models describing the dynamics of interacting species. They also allow computing a series of metrics associated with modern coexistence theory that inform about the likelihood of species to coexist. These metrics are 1) *niche differences* that stabilize coexistence between competing species and 2) *average fitness differences* that drive competitive dominance and, in the absence of niche differences, determine the superior competitor. 

The package also allows exploring how environmental variation modifies both the ability of species to produce offspring and the interaction coefficients between pairs of species (including with itself). This feature opens the possibility of exploring how stabilizing niche differences and average fitness differences vary across environmental gradients, and therefore, it allows analyzing whether the likelihood of species to coexist is constant across environmental conditions.  

Here we demonstrate the basic functionality of the package using a published observational dataset (see Lanuza et al. (2018) for a description of the dataset). With this example, we will specifically estimate seed production in the absence of neighbors (lambda) and the strength and sign of species interactions between species pairs (alpha matrix). These values are the basis for estimating the degree of niche overlap (i.e 1- niche differences) and average fitness differences between species pairs, which are covered in vignette 3.

**Fitting a single species**

First, we load the package and the associated data. The included dataset contains, for each individual, its reproductive sucess and the number of neighbors per species in a 7.5 cm buffer (see vignette 2).

```{r}
library(cxr)
data("neigh_list") 
```

First, we draw the values of a single focal species.

```{r}
my.sp <- "HOMA" 
# get data from the list
obs_homa <- neigh_list[[my.sp]]
# no need for ID column
obs_homa <- subset(obs_homa,select = -c(obs_ID))
# For each observation, we need the individual plant fitness and the number of neighbours per species (in columns).
head(obs_homa)
```

Next, we estimate both the reproduction success in the absence of neighbors (lambda) and the competition matrix (alpha). This is done by fitting a model that mathematically relates the reproductive success to the number of neighbors observed. In this first example, we fit the selected species with a Ricker model ('RK' model family, see vignette 4) and fairly standard initial values. The default optimization method (`Nelder-Mead`) does not allow for lower or upper bounds in model parameters, so these arguments are commented out. We can also specify whether we want to compute standard errors numerically, by setting the argument `bootstrap_samples` to the number of bootstrap replications for the calculation.

```{r}
#?cxr_pm_fit #check the help file for a description of the arguments
fit_homa <- cxr_pm_fit(data = obs_homa,
                       focal_column = my.sp,
                       model_family = "RK",
                       covariates = NULL,
                       # optimization_method = "Nelder-Mead",
                       alpha_form = "pairwise",
                       lambda_cov_form = "none",
                       alpha_cov_form = "none",
                       initial_values = list(lambda = 1,
                                             alpha_intra = .1,
                                             alpha_inter = .1),
                       # lower_bounds = list(lambda = 0,
                       #                     alpha_intra = 0,
                       #                     alpha_inter = 0),
                       # upper_bounds = list(lambda = 10,
                       #                     alpha_intra = 1,
                       #                     alpha_inter = 1),
                       fixed_terms = NULL,
                       bootstrap_samples = 3) # a low number for demonstration purposes, increase it for robust results.
```

For a quick summary of the fit, we can run a summary on the resulting object.

```{r}
summary(fit_homa)
```

The object is actually a list with several elements. We can thus access these elements as usual:

```{r}
names(fit_homa) #list of all available elements.

#reproduction success in the absence of neighbors
fit_homa$lambda
# intraspecific interaction
fit_homa$alpha_intra
# interspecific interactions
fit_homa$alpha_inter
```

**Fitting several species at once**

Most likely users will want to fit model parameters to data from two or more focal species. In order to do that with a single call, we provide the function `cxr_pm_multifit`, which has a very similar interface to `cxr_pm_fit`. Here we show how multiple species can be fit using this function. In order to showcase other capabilities of the package, we include in this example the effect of a covariate over the fitted lambda and alpha parameters. This covariate, soil salinity, is also included as a dataset in the package (see vignette 2).

```{r}
my.sp <- c("BEMA","CETE","LEMA")
obs_3sp <- neigh_list[my.sp]
# discard ID column
for(i in 1:length(obs_3sp)){
  obs_3sp[[i]] <- obs_3sp[[i]][,2:length(obs_3sp[[i]])]
}
# load covariates: salinity
data("salinity_list")
salinity <- salinity_list[my.sp]
# keep only salinity column
for(i in 1:length(salinity)){
  salinity[[i]] <- as.matrix(salinity[[i]][,2:length(salinity[[i]])])
  colnames(salinity[[i]]) <- "salinity"
}
```

Note how the data is passed in a list with as many elements as focal species. Each element is a dataframe with observations of the corresponding focal species. Same for the covariate, it must be a list with as many elements as focal species. Each element is a dataframe (or a matrix) with a single column and the same number of observations as its associated species data.

```{r}
names(obs_3sp)
head(obs_3sp[[1]])
nrow(obs_3sp[[1]])

head(salinity[[1]])
nrow(salinity[[1]])
```

And we fit the model as above, but using the `cxr_pm_multifit` function.

```{r}
fit_3sp <- cxr_pm_multifit(data = obs_3sp,
                           focal_column = my.sp,
                           model_family = "RK",
                           optimization_method = "L-BFGS-B", # use a bounded method for demonstration purposes
                           covariates = salinity,
                           alpha_form = "pairwise",
                           lambda_cov_form = "global", # effect of covariates over lambda
                           alpha_cov_form = "pairwise", # effect of covariates over alpha
                           initial_values = list(lambda = 1,
                                                 alpha_intra = 0.1,
                                                 alpha_inter = 0.1,
                                                 lambda_cov = 0.1,
                                                 alpha_cov = 0.1),
                           lower_bounds = list(lambda = 0,
                                               alpha_intra = 0,
                                               alpha_inter = -1,
                                               lambda_cov = 0,
                                               alpha_cov = 0),
                           upper_bounds = list(lambda = 100,
                                               alpha_intra = 1,
                                               alpha_inter = 1,
                                               lambda_cov = 1,
                                               alpha_cov = 1),
                           bootstrap_samples = 3)
```

We can also have a glimpse of this multispecies fit with the summary function:

```{r}
summary(fit_3sp)
```

The numerical estimation of parameters depends on the model with which to estimate fitness values, the optimization method, and the underlying data. In our example dataset, some species are better represented than others, and the function will raise warnings if the estimation can be improved or, for example, if any fitted parameter is equal to the lower or upper bound provided In the `cxr` objects returned from our functions, the log-likelihood of the fit is also included, which may be useful in helping users choose a certain optimization method for a particular dataset. In this example, the method used ("L-BFGS-B", a well-established bounded optimization algorithm) returns the following log-likelihood values:

```{r}
fit_3sp$log_likelihood
```

**Including environmental variability**

In the above example for fitting multiple species at once, we have already offered a glimpse on how to include the effect of environmental covariates over lambda and alpha values. The relevant arguments in `cxr_pm_fit` and `cxr_pm_multifit` are 'lambda_cov_form' and 'alpha_cov_form'. If these are set to 'none', no effect of covariates is considered. Otherwise, the effect of covariates over lambda can be 'global', meaning that each covariate has a global parameter affecting lambda. This is formulated as

$\lambda (1 + \sum_{k=1}^{s} \theta_k c_k)$

where $s$ is the number of environmental covariates, $c_k$ is the observed value of the i-th covariate and $\theta_k$ is the 'lambda_cov' parameter of the `cxr` functions.

The effect over alpha values can likewise be 'global', so that, for focal species $i$, each covariate affects the alpha values $\alpha_{ij}$ equally through a global parameter:

$\alpha_* + \sum_{k=1}^{s} \psi_k c_k$

where $*$ represents the set of all pairwise interaction for a given focal species, and $\psi_k$ is the global parameter relating covariate $k$ to the alpha values. The last possibility included in `cxr` is for the effect of covariates over alpha values to be interaction-specific, which is coded by specifying 'alpha_cov_form' as 'pairwise':

$\alpha_{ij} + \sum_{k=1}^{s} \psi_{ij} c_k$

In this case, each covariate $c_k$ will have a specific parameter $\psi_{ij}$ for its effect over each interaction $\alpha_{ij}$.

We can retrieve the fitted values for 'lambda_cov' ($\theta$) and 'alpha_cov' ($\psi$) simply from the output of the function:

```{r}
fit_3sp$lambda_cov
fit_3sp$alpha_cov
```

**References**

Lanuza, J. B., Bartomeus, I., & Godoy, O. (2018). Opposing effects of floral visitors and soil conditions on the determinants of competitive outcomes maintain species diversity in heterogeneous landscapes. Ecology letters, 21(6), 865-874.


