---
title: "Data formats"
output: rmarkdown::html_vignette
author: David García-Callejas and cxr team
vignette: >
  %\VignetteIndexEntry{Data and model formats}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

**Introduction**

The `cxr` package provides diverse functions to handle empirical datasets, yet there are limitations in data format. Here we present the structure of the dataset included in this package, which is in the specific formats accepted according to the developed functions. 

**The Caracoles dataset**

We include a dataset of an annual plant system subjected to spatial variability in a Mediterranean-type ecosystem of South Europe. The sampling design is explained in *Lanuza, Bartomeus, Godoy, 2018, Ecology Letters 21:865-874*. The main data file contains, for each individual sampled, its reproductive sucess and the number of neighbors per species in a 7.5 cm buffer. Note that this format of data is not limited to plant species, but can contain any information of individual performance as a function of the interacting species density. 

```{r}
library(cxr)
data("neigh_list", package = "cxr")
```

You can check the structure of the data in the help file

```{r}
?neigh_list
names(neigh_list)
head(neigh_list[[1]])
```

For each sub-plot, we also recorded the species abundance:

```{r}
data("abundance", package = "cxr")
head(abundance)
```

We also provide seed soil survival and germination rates for each species. These species vital rates have been obtained independently. They are critical to parameterize a model describing the population dynamics of interacting annual plant species. This file also includes the complete scientific name and abbreviation of each species. Such abreviation will be used as species identifier in subsequent analyses and vignettes.

```{r}
data("species_rates", package = "cxr")
species_rates ##SPECIES COMBINA MINUSCULAS Y MAYUSCULAS; MOLARIA UNIFICAR A Gen sp.
```

The environmental covariate provided for this analysis is soil salinity measured with a portable TDR (i.e. Time Domain Reflectometer). This technology measures the amount of salt disolved in the soil water that is accesible to plant species. This environmenal covariate has been estimated for each sub-plot: 

```{r}
data("salinity_list", package = "cxr")
names(salinity_list)
head(salinity_list[[1]])
```
